Static Analysis by Path Focusing

### 2011-05-16 report/report.tex
Bug fix: there was a bug in the case of OR, AND,... instructions for integer variables

\usepackage{mathcomp,amssymb,verimagstudent}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyzer, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of the these two methods.
\end{itemize}

@@

Static analysis aims to discover some properties about programs. In this
example, one would to compute the set of possible values for the variable $x$
during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

@@

This computation is a fixpoint computation: one updates the different sets until
we find no more elements. These sets are strictly increasing (we always add new
elements in the sets), and if we find no new elements to add, that means we have
reached a limit. Abstract Interpretation gives a general framework to do such
kind of fixpoint computation.

@@

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& \Phi^\#(x_n)
\end{array}
\right.
$$
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. There have been various
work proposing a definition of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against the loss of precision induced by widening.

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, one could choose $P_w$ as the set of the
heads of loops.

@@

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to various reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, one compute a union of convex polyhedra instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, one obtain $i \in
[0,1]$, \dots. After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better. We can repeat such iterations as long as the set narrows, and
we always obtain a correct over-approximation.
In practice, a single narrowing iteration is enough to recover the precision.

@@

To limit the number of such points, a method could be
to expand the control flow graph. For instance, instead of considering a graph
with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, one could remove the $n$ merge node
of these \emph{if-then-else}, each of them having two incoming edges, and adding 
$2^n$ edges from $p_1$ to $p_n$, corresponding to the different paths through
the \emph{if-then-else} statements. At the end, $p_n$ is the only point with
several incoming edges. This graph transformation results in an exponential
blowup, because of the exponential growth of the number of edges in the graph.
\end{enumerate}

@@

Typically, to analyze a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, one could directly compute an invariant for the loop:

@@

Lookahead widening \cite{GopanR06} is a technique that isolates the
different loop phases during the analysis, and stabilizes each of these loop
phases before proceeding to the next. One only consider paths that are feasible
at the first iteration, and forget the others for a while. The iterations end
when the analysis of these paths has converged. After that, one can compute some
narrowing iterations. Then, paths that
become feasible are reinserted into the control flow graph, and we redo
iterations.

@@

	Let $(P,E)$ be the control flow graph of the program.
	First, one chooses a set of widening points $P_W \subseteq P$, so that the graph
	obtained after removing these points has no cycle. The widening operator
	will be applied at these points. The choice of $P_W$ can be guided by
	\cite{Bou92}.
	Then, one chooses another set $P_R \subseteq P$ of nodes, satisfying the
	following properties:
	\begin{itemize}
	\item $I_p = \emptyset$ for each $p \in P \setminus P_R$, meaning that the initial
	points of the program are included in $P_R$.
	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
	\end{itemize}

@@

	This is a reachability problem, that can be expressed as an SMT-formula,
	which is a logic formula with Boolean variables, and elements of a certain
	theory $T$. Depending on the program, one can choose different theories:
	\begin{itemize}
	\item For programs with rational variables, whose operations (instructions,
	transition conditions\dots) are all linear arithmetic, $T$ can be the theory
	of linear real arithmetic (LRA).
	\item For programs with integer variables (with a numerical state space in
	$\Z^n$) $T$ can be the theory of linear integer arithmetic (LIA).
	\end{itemize}

@@

If the point where the $\Phi$-variable $x$ is defined is in
$P_R$, one need to rename $x$ into $x'$:
\begin{itemize}
\item $x$ will be the value of $x$ at the beginning of the path.
\item $x'$ will be the value of $x$ at the end of the path.
\end{itemize}
Indeed, there can be uses of $x$ along the path, and these uses refer to the
old value of $x$. 
\end{enumerate}

@@

	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
	We maintain a set $A$ of point in $P_R$ that need to be treated.
	We select iteratively one of these control points, and search for new focus
	paths starting from this point, until there is no more path to be treated.
	At the end of the analysis, one could perform some narrowing iterations in
	order to recover some precision.
\begin{algorithm}
\caption{Path Focusing}\label{pathfocusingalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\State $Y \gets \tau_e^\#(X_{p_i})$
		\If {$p_j \in P_W$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly perform some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply apply the widening
	operator, one could compute some narrowing steps to regain precision before
	focusing on another path. This is the same principle as the one we find in
	\emph{Lookahead widening} \cite{GopanR06}.

@@

	The $SelfLoop$ procedure actually computes a widening operation, and
	computes a narrowing step in order to recover precision. In some cases,
	instead of performing widening/narrowing iterations, one could use
	acceleration techniques to find even more precise invariants \cite{GH06}.

	In order for this refinement to be precise, one needs to delay the widening
	operator when we consider a self-loop for the first time. Then, our
	algorithm uses a set $U$ of paths that have already been treated. If the
	self-loop we focus on is not in $U$, we apply a simple union instead of
	widening. Widening will be applied next time to guarantee the convergence.

	When the self loop is accelerable \cite{Gon07}, instead of doing a
	widening/narrowing sequence, one can directly accelerate the loop and find
	its transitive closure.

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. One computes
	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.

@@

This algorithm requires to enumerate all the paths and to compute their
semantics, which may result in an exponential blowup. In addition, the
\emph{for} loop at line $7$ iterates on all the $n \times m$ values of $(i,j)$:
some of the join operations may be useless, in the sense that the value of 
$X_{p,\sigma(j,i)}$ may not change. Here, one could use the same technique as
Path Focusing to avoid these drawbacks: one can use SMT-solving to find the
paths $\tau_{p,i}$.

@@

@@

LLVM internal representation can be applied transformation passes that modifies
parts of it. These passes are mostly used for optimisation purpose. In our
analyzer, we apply some already existing passes before running our 
abstract interpretation pass. Among the passes we run, one could cite:
\begin{itemize}
\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
representation by promoting the memory variables to Static Single Assignment
(SSA) registers. This pass transform the control flow graph such that all the
variables are in SSA form, and removes most of the \emph{Load/Store} operations
into memory. It is useful since our analyzer doesn't have a memory model, and
then loses lots of precision in the case of \emph{Load/Store} instructions.
\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
format, and compute some information related to loops (for instance, the set of
BasicBlocks that are parts of the loop, etc.).
\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
corresponding \emph{if-then-else} instructions. After running this pass, there
is no more \emph{Switch} instruction in the control flow graph, so the
implementation of the static analysis is easier.
\end{itemize}

@@

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is inexistant in the representation, since the compiler doesn't
need this information to generate code. Yet, for static analysis, one need to
have this distinction to improve precision:
for instance, if $n$ is unsigned, one could add the constraint $n \geq 0$ in our
	abstract value at the beginning of the analysis.
	This lack of precision may be limited by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

@@

To do so, one could implement an LLVM pass that transform the internal
representation according to this principle. However, this pass has not been
implemented in our analyzer, and this transformation has to be done manually.

@@

@@

	In this section, we evaluate our \emph{Lookahead Widening} and \emph{Path
	Focusing} implementations. We can compare their execution time in some
	practical cases, and the precision of their results. Table \ref{experiments}
	gives the results of the analyzer in various examples, some of them taken
	from the recent literature \cite{CGGMP05,GopanR06}.

@@

>>>>

\usepackage{mathcomp,amssymb,verimagstudent}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[toc,page]{appendix}

@@

In this report we make the following contributions:
\begin{itemize}
\item We present a new technique, refered as \emph{Path Focusing}, that takes
benefits of the recent advances in the field of SMT-solving to improve precision
of static analysis by abstract interpretation. We also present an adaptation of
this technique to compute disjunctive invariants.
\item We present an implementation of this technique into a small analyzer, as
well as the implementation of another technique proposed in \cite{GopanR06},
such that we can compare precision and cost of these two methods.
\end{itemize}

@@

Static analysis aims to discover some properties about programs. In this
example, one would compute the set of possible values for the variable $x$
during execution.
The graph at the right side represents the program. The nodes of the graph are
called \emph{program points}. We can compute for instance the set $Y_1$ of
values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
the second one from $p_2$.

@@

This computation is a fixpoint computation: the different sets are updated until
we find no more elements. These sets are strictly increasing (we always add new
elements in the sets), and if we find no new elements to add, that means we have
reached a limit. Abstract Interpretation gives a general framework to do such
kind of fixpoint computation.

@@

$$\left[
\begin{array}{lll}
x_0 &=& \perp \\
x_{n+1} &=& \Phi^\#(x_n)
\end{array}
\right.
$$

where $\perp$ is the least element of $A$.
%The relation between these two
%domains is characterized by two functions, usually called 
%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
%
%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
%\leq_{C} \gamma(y) $$
%
%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
%fixpoint computation is done in the abstract domain, using the approximation of
%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
%noted $\Phi^\#$.
%We have the following result:
%
%\begin{proposition}
%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
%\end{proposition}
%
%In fact, one compute the least fixpoint of the function in the abstract
%domain, and the result of the computation also gives an upper approximation of
%the fixpoint in the concrete domain.

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. 
Intuitively, the polyhedron $P \nabla Q$ is obtained after 
removing from the system of linear equations defining $P$ all the inequalities
that are not satisfied by $Q$.
For instance, if $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$ and $Q = 
\{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$, then the result of the widening operator
will be $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$.

There have been various
work proposing a definition or a refinment of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against loss of precision:
\begin{itemize}
\item \emph{Delayed widening}: instead of applying windening at the first
iteration of the analysis, one apply it after a number $n$ of iterations. During
the $n-1$ first iterations, only a convex hull is computed. In some cases,
delaying the widening operation is more precise. The algorithm still terminates,
since widening is applied within a finite time after $n$ iterations. Most of the
cases, $n$ can be choosen equals to $2$.
\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
brutally applying the widening operator, a fixed set of linear inequalities $T$
can be chosen, and the widening operator $\nabla_T$ is defined as follows:
$P \nabla_T Q$ is the intersection of $P \nabla Q$ with all the inequalities in
$T$ satisfied by both $P$ and $Q$.

For instance, consider the program:
\begin{C}
while (x <= 50) {
	x++;
	y++;
}
\end{C}
If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
widening (which gives the result $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
a widening with the threshold $x \leq 50$ gives 
$P \nabla_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
with threshold helps to find better invariants.
\end{itemize}

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
heads of loops.

@@

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to various reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, a union of convex polyhedra is computed instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, $i \in
[0,1]$, then $i\in [0,2]$ after the second iteration\dots. 
After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better. We can repeat such iterations as long as the set narrows, and
we always obtain a correct over-approximation.
In practice, a single narrowing iteration is enough to recover the precision.

@@

To limit the number of such points, a method could be
to expand the control flow graph. For instance, instead of considering a graph
with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, 
the $n$ merge node
of these \emph{if-then-else} could be removed, each of them having two incoming
edges, and 
$2^n$ edges from $p_1$ to $p_n$ are added, corresponding to the different paths through
the \emph{if-then-else} statements. At the end, $p_n$ is the only point with
several incoming edges. This graph transformation results in an exponential
blowup, because of the exponential growth of the number of edges in the graph.
\end{enumerate}

@@

Typically, to analyze a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, an invariant could be directly computed for the loop:

@@

Lookahead widening \cite{GopanR06} is a technique that isolates the
different loop phases during the analysis, and stabilizes each of these loop
phases before proceeding to the next. One only consider paths that are feasible
at the first iteration, and forget the others for a while. The iterations end
when the analysis of these paths has converged. After that, some
narrowing iterations can be computed. Then, paths that
become feasible are reinserted into the control flow graph, and we redo
iterations.

@@

	Let $(P,E)$ be the control flow graph of the program.
	First, one chooses a set of widening points $P_W \subseteq P$, so that the graph
	obtained after removing these points has no cycle. The widening operator
	will be applied at these points. The choice of $P_W$ can be guided by
	\cite{Bou92}.
	Then, another set $P_R \subseteq P$ of nodes is chosen, satisfying the
	following properties:
	\begin{itemize}
	\item $I_p = \emptyset$ for each $p \in P \setminus P_R$, meaning that the initial
	points of the program are included in $P_R$.
	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
	\end{itemize}

@@

	This is a reachability problem, that can be expressed as an SMT-formula,
	which is a logic formula with Boolean variables, and elements of a certain
	theory $T$. Depending on the program, different theories may be used:
	\begin{itemize}
	\item For programs with rational variables, whose operations (instructions,
	transition conditions\dots) are all linear arithmetic, $T$ can be the theory
	of linear real arithmetic (LRA).
	\item For programs with integer variables (with a numerical state space in
	$\Z^n$) $T$ can be the theory of linear integer arithmetic (LIA).
	\end{itemize}

@@

If the point where the $\Phi$-variable $x$ is defined is in
$P_R$, $x$ needs to be renamed into $x'$:
\begin{itemize}
\item $x$ will be the value of $x$ at the beginning of the path.
\item $x'$ will be the value of $x$ at the end of the path.
\end{itemize}
Indeed, there can be uses of $x$ along the path, and these uses refer to the
old value of $x$. 
\end{enumerate}

@@

	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
	We maintain a set $A$ of point in $P_R$ that need to be treated.
	We select iteratively one of these control points, and search for new focus
	paths starting from this point, until there is no more path to be treated.
	At the end of the analysis, some narrowing iterations can be performed in
	order to recover some precision.
\begin{algorithm}
\caption{Path Focusing}\label{pathfocusingalgo}
\begin{algorithmic}[1] 
\Procedure{pathfocusing}{$P$,$E$,$P_R$}
\ForAll {$p \in P_R$}
	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
\EndFor
\State $\rho \gets$ computeRho$(P,E)$
\State $A \gets \emptyset$
\ForAll {$p \in P_R / I_p \neq \emptyset$}
	\State $A \gets A \cup p$
\EndFor
\While{$A \neq \emptyset$}
	\State Select $p_i \in A$
	\State $A \gets A \setminus \{p_i\}$
	\While{true}
		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
		\displaystyle\bigwedge_{\substack{j\neq i \\
		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
		X_{p_j})\right)\right]$
		\If {$res = unsat$}
			\State \textbf{break}
		\EndIf
		\State Compute the focus path $e$ from $p_i$ to $p_j$
		\State $Y \gets \tau_e^\#(X_{p_i})$
		\If {$p_j \in P_W$}
			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
		\Else
			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
		\EndIf
		\State $A \gets A \cup \{p_j\}$
	\EndWhile
\EndWhile
\State Possibly perform some narrowing steps
\State Compute $\{X_{p_i},\ i \notin P_R\}$
\State \textbf{return} $\{X_{p_i},\ i \in P\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply apply the widening
	operator,  some narrowing steps could be performed to regain precision
	before focusing on another path. This is the same principle as the one we
	find in \emph{Lookahead widening} \cite{GopanR06}.

@@

	The $SelfLoop$ procedure actually computes a widening operation, and
	computes a narrowing step in order to recover precision. In some cases,
	instead of performing widening/narrowing iterations, 
	acceleration techniques could be used 
	to find even more precise invariants \cite{GH06}.

	In order for this refinement to be precise, the widening
	operator needs to be delayed
	when we consider a self-loop for the first time. Then, our
	algorithm uses a set $U$ of paths that have already been treated. If the
	self-loop we focus on is not in $U$, we apply a simple union instead of
	widening. Widening will be applied next time to guarantee the convergence.

	When the self loop is accelerable \cite{Gon07}, instead of doing a
	widening/narrowing sequence,  acceleration directly finds
	its transitive closure.

@@

	We apply the \emph{Path Focusing} technique to our running example from
	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
	\ref{multigraphGopan}.
	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
	The $Ŕho$ formula associated to this program is depicted in
	Figure \ref{rhoformula}.
	One computes
	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.

@@

This algorithm requires to enumerate all the paths and to compute their
semantics, which may result in an exponential blowup. In addition, the
\emph{for} loop at line $7$ iterates on all the $n \times m$ values of $(i,j)$:
some of the join operations may be useless, in the sense that the value of 
$X_{p,\sigma(j,i)}$ may not change. Here, the same technique as
Path Focusing is achievable to avoid these drawbacks:  SMT-solving is used
to find the paths $\tau_{p,i}$.

@@

	\section{Efficiency comparison: example}

In this subsection, we propose an example showing that \emph{Path Focusing}
technique performs better than \emph{Lookahead Widening} in the case of loops
with intermediate control-flow merges.

We consider the program depicted in figure \ref{ex9}.
Here, we would like to show that the returned value $r$ is zero.

\begin{figure}[!h]
\begin{minipage}[c]{.49\linewidth}
\begin{C}
x = 0;
i = 0;
r = 0;
while (i < 100) {
	if (i < 50)
		x = -1;
	else
		x = 1;

	if (x == 0) r++;
	i++;
}
return r;
\end{C}
\end{minipage}
\begin{minipage}[c]{.49\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
                    semithick,font=\footnotesize]

	\node[state] (n0) {$p_0$};
	\node[state] (n1) [below of=n0] {$p_1$};
	\node[state] (n2) [below of=n1] {$p_2$};
	\node[state] (n3) [below left of=n2] {$p_3$};
	\node[state] (n4) [below right of=n2] {$p_4$};
	\node[state] (n5) [below right of=n3] {$p_5$};
	\node[state] (n6) [below left of=n5] {$p_6$};
	\node[state] (n7) [below right of=n5] {$p_7$};
	\node[state] (n8) [below left of=n7] {$p_8$};
	\node[state] (n9) [left of=n1] {$p_9$};

  \path [transition] 
		(n0) edge              node {$x,i,r \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge              node {$i < 100$} (n2);
  \path [transition] 
		(n1) edge              node {$i \geq 100$} (n9);
  \path [transition] 
        (n2) edge			   node [left] {$i < 50$} (n3);
  \path [transition] 
        (n2)  edge              node [right] {$i \geq 50$} (n4);
  \path [transition] 
        (n3) edge              node [left] {$x \leftarrow -1$} (n5);
  \path [transition] 
        (n4) edge			   node [right] {$x \leftarrow 1$} (n5);
  \path [transition] 
        (n5) edge			   node [left] {$x = 0$} (n6);
  \path [transition] 
        (n5) edge			   node [right] {$x \neq 0$} (n7);
  \path [transition] 
        (n6) edge              node [left] {$r \gets r+1$} (n8);
  \path [transition] 
        (n7) edge              node {} (n8);
  \path [transition] 
        (n8) edge [out=0, in=0, distance=3.5cm] node [right] {$i \gets i+1$} (n1);

\end{tikzpicture}
\end{minipage}
\caption{Example of program. $p_5$ is a control-flow merge point, which is in
the middle of the containing loop.}
\label{ex9}
\end{figure}

\paragraph{Lookahead Widening} 
The analysis begins with $p_1: x=i=r=0$. Then, the path $p_1 \rightarrow p_2
\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$ is the only one
possible. The technique is then to stabilize the iterations over this subgraph,
removing $p_4$ and $p_6$ for a while. After stabilization and narrowing phases,
we obtain $ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at
control point $p_1$, and 
$ 0 \leq i \leq 49 \wedge r=0 \wedge x=-1$ at point $p_5$.

Then, point $p_4$ becomes reachable. We continue the iterations and have to do
a convex hull of polyhedra at point $p_5$, because of the two possible incoming
transitions.
We obtain the polyhedron
$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 1$.
Point $p_6$ now becomes reachable, since $x=0$ is inside our polyhedron.
Finally, the analysis terminates after some steps and finds $0 \leq r \leq i$.
\emph{Lookahead Widening} is not able to prove that $r = 0$ at the end of the
funtion.

\paragraph{Path Focusing} 
The analysis of the loop starts with 
$p_1: x=i=r=0$. Then, the SMT-solver discovers the path $p_1 \rightarrow p_2
\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. We do a
widening/narrowing sequence and we find 
$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The smt gives
another path: 
$p_1 \rightarrow p_2
\rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
analyzing this path, we obtain 
$ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
SMT-solver gives $unsat$ answers. Here, We never focused on paths going through
point $p_6$, so our analysis gives us the invariant $x=0$ at the end of the
program. Indeed, there is no feasible paths going through $p_6$, but
\emph{Lookahead Widening} don't know it because of the convex hull at point
$p_5$.

@@

LLVM internal representation can be applied transformation passes that modifies
parts of it. These passes are mostly used for optimisation purpose. In our
analyzer, we apply some already existing passes before running our 
abstract interpretation pass. Among the passes we run, we could cite:
\begin{itemize}
\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
representation by promoting the memory variables to Static Single Assignment
(SSA) registers. This pass transform the control flow graph such that all the
variables are in SSA form, and removes most of the \emph{Load/Store} operations
into memory. It is useful since our analyzer doesn't have a memory model, and
then loses lots of precision in the case of \emph{Load/Store} instructions.
\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
format, and compute some information related to loops (for instance, the set of
BasicBlocks that are parts of the loop, etc.).
\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
corresponding \emph{if-then-else} instructions. After running this pass, there
is no more \emph{Switch} instruction in the control flow graph, so the
implementation of the static analysis is easier.
\end{itemize}

@@

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is inexistant in the representation, since the compiler doesn't
need this information to generate code. Yet, for static analysis, 
this distinction is required to improve precision:
for instance, if $n$ is unsigned, the constraint $n \geq 0$ could be added in
	our abstract value at the beginning of the analysis.
	This lack of precision may be limited by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

@@

To do so, we could implement an LLVM pass that transform the internal
representation according to this principle. However, this pass has not been
implemented in our analyzer, and this transformation has to be done manually.

@@

Additionally, some optimization in the code could be proposed in order to
reduce time and/or memory complexity. For instance, paths that have already been
computed are until now stored in a tree. An optimization could be to store them
in a Binary Decision Diagram, in order to reduce significantly the memory
consumption when there is a huge number of paths.

@@

	In this section, we evaluate our \emph{Lookahead Widening} and \emph{Path
	Focusing} implementations. We compare their execution time in some
	practical cases, and the precision of their results. Table \ref{experiments}
	gives the results of the analyzer in various examples, some of them taken
	from the recent literature \cite{CGGMP05,GopanR06}.

@@

\emph{Path Focusing} appears to always perform similarly or better than 
\emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
containing control-flow merges inside of loops, because of the distinction
between all the paths of the loop. In simpler programs, it performs as well as
\emph{Lookahead Widening}

\emph{ex6.c} is a program containing non-linear arithmetic: there is an integer
multiplication inside the loop. The Apron library handles multiplication using
a linearization technique. However, multiplications is not allowed in LIA
(Linear Integer Arithmetic) or LRA (Linear Real Arithmetic). Then, Yices
SMT-solver errs with a message ``not implemented'' (which is the reason of the
$\infty$ symbol in the table), when Z3 still works correctly.

We tried our technique on example with a huge number of paths: ex8, ex8b and
ex8c (see Table \ref{npaths}). These benchmarks are constructed such that the
loop contains only a few
possible paths, but a lot of unfeasible paths. The idea is to show that our
technique doesn't blow up, even with a huge expanded multigraph. Results show
that in such benchmarks, our technique finds a much better invariant and is even
faster than \emph{Lookahead Widening}.

For small programs, our execution time may not be significant. Indeed, the cost
for reading the program, initializing the SMT-solver, \dots, may have a
	non-negligible cost compared to the abstract interpretation analysis itself.
	The only interest of these times is to show our algorithm is pretty fast.
